---
title: "Code for Slides 3/28: Data Joins"
author: "Grayson White"
date: "March 28, 2023"
---

## Example Data

For the following examples of data joins, we will use the tibbles from the first slide. We can load this data into `R`:

```{r, echo = TRUE}
library(tidyverse)
x <- tibble(spp_code = c("ABBA", "BEPA", "PIST"),
            common_name = c("balsum fir", "paper birch", "white pine"))

y <- tibble(spp_code = c("ABBA", "BEPA", "LALA"),
            latin_name = c("Abies balsamea", 
                           "Betula papyrifera",
                           "Larix laricina"))
```

## Mutating Joins

`dplyr` contains four mutating joins: 

1. `left_join(x, y)` keeps all rows of `x`, but if a row in `y` does not match to `x`, an `NA` is assigned to that row in the new columns. 
2. `right_join(x, y)` is equivalent to `left_join(y, x)`, except for column order.  
3. `inner_join(x, y)` keeps only the rows matched between `x` and `y`.
4. `full_join(x, y)` keeps all rows of both `x` and `y`. 

## Examples: Mutating Joins

Recall our example data

```{r, echo = TRUE}
x
y
```

## `left_join()`

```{r, echo = TRUE, message = TRUE}
left_join(x, y)
```

But what's that message about? 

    "Joining with `by = join_by(spp_code)`"

## We need to specify a *key* 

A *key* is can just be thought of the name(s) of the column(s) you're joining by. In the `left_join()` from the last slide, `R` assumed we were joining by the column `spp_code` since both `x` and `y` have a column with that name. 

Keys are important, especially when the columns you are joining by have different names, or you are joining by multiple columns. 


## `left_join()`, with a 

```{r, echo = TRUE, message = TRUE}
left_join(x, y, by = "spp_code")
```

-   Notice that we specify this key with the `by` argument. This is the same for all joins in `dplyr`. 

## `right_join()`

```{r, echo = TRUE, message = TRUE}
right_join(y, x, by = "spp_code")
```

Notice that this is the same as our previous `left_join()`.

What happens if we try switching the order of `x` and `y`?

```{r, echo = TRUE}
right_join(x, y, by = "spp_code")
```

## `inner_join()`

How many rows will the output have?

```{r, echo = TRUE}
inner_join(x, y, by = "spp_code")
```

Why is this the result? 

## `full_join()`

How many rows will the output have?

```{r, echo = TRUE}
full_join(x, y, by = "spp_code")
```

Why is this the result?

## Filtering Joins

`dplyr` contains two filtering joins: 

1. `semi_join(x, y)` keeps all the rows in `x` that have a match in `y`. 
2. `anti_join(x, y)` removes all the rows in `x` that have a match in `y`.

**Note**: Unlike mutating joins, filtering joins do not add any columns to the data. 

## `semi_join()`

How many rows will this `semi_join` return? How many columns?

```{r, echo = TRUE}
semi_join(x, y, by = "spp_code")
```

## `semi_join()`

What about this `semi_join`? Will it be the same as `semi_join(x, y)`? 

```{r, echo = TRUE}
semi_join(y, x, by = "spp_code")
```

## `anti_join()`

Let's see what `anti_join` does:

```{r, echo = TRUE, output = T}
anti_join(x, y, by = "spp_code")
```

Why do we get this output? 

## `anti_join()`

What happens if we switch the order of `x` and `y`? 

```{r, echo = TRUE, output = T}
anti_join(y, x, by = "spp_code")
```

## An Important Subtlety: Column Names

So far, we have joined `x` and `y` by the `spp_code` column. 

But what if `y` had the same column named differently:

```{r, echo=T, output=T}
y <- y %>%
  rename(species_code = spp_code)
y
```

## How Do We Join `x` and `y`? 

```{r, output=T, echo=T, error = T}
left_join(x, y)
```

Looks like we need to specify `by` (our key)

```{r, output=T, echo=T, error = T}
left_join(x, y, by = "spp_code")
```

Still not working!

```{r, output=T, echo=T}
left_join(x, y, by = c("spp_code" = "species_code"))
```

